---
title: "chromosome_level_mapping.Rmd"
author: "Emily Dombrowski"
date: "2026-07-07"
output: html_document
---
This is a script for analyzing genome level trends in my mussel samples. I have adult and larval samples. To make the plots, I am using outputs from the package 'samtools' and heavily consulting (r graph gallery)[https://r-graph-gallery.com/]. I have described the data in more detail below and the processes of analyzing. 


#### Load libraries and files
I am going to be graphing with ggplot and use plotly to visualize sample level data.
The file that I am loading contains coverage data per chromosome/scaffold for each of the samples. 
```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(plotly)

# samtools coverage data outputs
cov_data <- read_delim("coverage_chrom.txt", delim = "\t")
# picard mapping proportions outputs
qc_data <- read_delim("mapping_proportions_summary.txt", delim = "\t")

head(cov_data)
```
Some notes about each row: 
* sample: sample name, in the format "Sample ID_temperature_day_elution"; some deviations from this
* chromosome: chromosome name
* start_pos: first place read on chromosome (always 1)
* end_pos: total length of chromosome
* numreads: total amount of reads that aligned to this chromosome
* covbases: number of base pairs along chromosome that have at least one read per site
* coverage: percentage of chromosome's length that was physicaly sequenced
* meandepth: number of reads stacked vertically per chromosome
* meanbaseq: mean base quality
* meanmapq: mean map quality

Next, I am going to do some manipulations to assign filters to each variable that separate them by temperature and life stage.
Additionally, there are some names that don't follow the strict pattern, so I included code to manually reassign these samples. 

#### manipulate data to add filters
```{r}
cov_filter <- cov_data %>%
  mutate(timepoint = case_when(
      # Looks for strings starting with F or M
      str_detect(sample, "^[FM]")       ~ "adult",
      
      # Matches exact day tags within the sample ID
      str_detect(sample, "-D1-")          ~ "day_1",
      str_detect(sample, "-D10-")         ~ "day_10",
      
      # The | operator means "OR", catching both Day 19 and Day 20
      str_detect(sample, "-D19-|-D20-")   ~ "final_day",
      
      # Catch-all safety net just in case a sample ID is misformatted
      TRUE                                ~ "Unknown" 
    ),
    
    # 2. Extract the Temperature component
    temperature = case_when(
      str_detect(sample, "-14C-") ~ "14C",
      str_detect(sample, "-22C-") ~ "22C",
      TRUE                        ~ "initial" # Adults and Day 1 baselines land here
    )
    
  )
  
cov_filter <- cov_filter %>%
  mutate(
    temperature = case_when(
      sample == "C3R3-D20-14-E1" ~ "14C",
      sample == "B1R3-B2R1-22-D10-E1" ~ "22C",
      TRUE ~ temperature
                          )
        )

# head(cov_filter)

qc_data <- qc_data %>%
  mutate(
    # The original 'Total_Reads' column actually represents the true mapped reads
    True_Mapped = Total_Reads, 
    True_Total  = Total_Reads + Unmapped_Reads,
    
    # Recalculate everything accurately
    Proportion_Mapped = True_Mapped / True_Total,
    Total_Reads       = True_Total,
    Mapped_Reads      = True_Mapped
  ) %>%
  dplyr::select(-True_Mapped, -True_Total)
```
Now, the timepoint and temperature columns can be used for downstream filtering. Let's start calculating coverage, reads, etc. per sample and chromosome. 

Some configuring from picard data outputs to join data frames: 

```{r}
# get rid of groups to make this join more straight forward
metadata_df <- cov_filter %>%
  group_by(sample, timepoint, temperature) %>%
  summarize(.groups = "drop") 

mapping_proportions <- metadata_df %>%
  left_join(qc_data, by = c("sample" = "Sample"))

head(mapping_proportions)

```

```{r}
# convert temperature to a factor to have graph better sorted
mapping_proportions$temperature <- factor(mapping_proportions$temperature, 
                                          levels = c("initial", "14C", "22C"))

# Barplot showing sample vs. proportion mapped
p <- ggplot(mapping_proportions, aes(x = sample, y = Proportion_Mapped, fill = temperature)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank()) +
  labs(y = "Mapped reads / total reads",
       x = "Sample (hover over bar to see individuals)")
p

ggplotly(p)

# Total reads per sample
m <- ggplot(mapping_proportions, aes(x=sample, y=Mapped_Reads)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank()) +
  labs(y = "Mapped reads",
       x = "Sample")

ggplotly(m)

box <- ggplot(mapping_proportions, aes(x = timepoint, y = Proportion_Mapped, fill = temperature)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(text = sample), position = position_jitterdodge(jitter.width = 0.1), size = 2) +
  theme_minimal() +
  labs(x = "Life stage", y = "Mapped reads/unmapped reads")

ggplotly(box)

print(box)
```



I think I want a plot with an x axis that is "chromosome," and a y axis that is "sample coverage." There will be 70 points per chromosome...I wonder if it will be informative to break it up by temperature.

#### coverage accross the genome by temperature treatment
```{r coverage accross the genome by treatment}

cov_chrom <- ggplot(cov_filter, aes(x = chromosome, y = coverage, fill = temperature, color = temperature)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, position = position_dodge(width = 1.0)) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
              size = 1.5, alpha = 0.5, aes(text = sample)) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(x = "Chromosome", y = "Coverage")

print(cov_chrom)

ggplotly(cov_chrom)
```



Okay, so everything looks pretty high. I am happy with this! But, this graph is kind of annoying to look at, so let's do a more traditional plot. 
x-axis: chromosome
y axis: coverage
Plot type: points

#### coverage accross the genome
```{r coverage accross the genome}

cov_gen <- ggplot(cov_filter, aes(x = chromosome, y = coverage, fill = temperature)) +
  geom_point() +
  theme(axis.text.x = element_blank()) +
  theme_minimal() +
  labs(x = "Chromosome", y = "Coverage")

print(cov_gen)

ggplotly(cov_gen)

```

Okay, now coverage by sample. I think to do this, we need to add a new feature to our df that compresses the average coverage per sample/the coverage per sample. I might also try and plot depth on the same plot...could be cool to see the two together. I will have to make sure to not show the sample names because it will be quite a lot all at the same time. 

#### coverage and depth by sample
```{r}

sample_qc <- cov_filter %>%
  group_by(sample, timepoint, temperature) %>%
  summarize(
    # calculate sample coverage using the end position and coverage per sample
    global_coverage = (sum(covbases)/sum(end_pos)) * 100,

    glob_depth = sum(meandepth * end_pos) / sum(end_pos),
    .groups = "drop"
  )
  
# cov_samp <- ggplot(sample_qc, x = sample, y = global_coverage) +
#  geom_point(fill = temperature) +
#  theme(axis.text.x = element_blank())


cov_samp <- ggplot(sample_qc, aes(x = sample, y = global_coverage, color = temperature)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    x = "Sample",
    y = "Coverage (%)",
    color = "Temperature"
  ) +
  theme(
    axis.text.x = element_blank(), 
    panel.grid.minor = element_blank()
  )

print(cov_samp)

cov_box <- ggplot(sample_qc, aes(x = timepoint, y = global_coverage, fill = temperature)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(text = sample), position = position_jitterdodge(jitter.width = 0.1), size = 2) +
  theme_minimal() +
  labs(x = "Life stage", y = "Coverage")

print(cov_box)

ggplotly(cov_box)

#### global genome depth by sample

depth <- ggplot(sample_qc, aes(x = sample, y = glob_depth, color = temperature)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    x = "Samples",
    y = "Depth",
    color = "Temperature") +
  theme(axis.text.x = element_blank())

print(depth)

#### average sample depth accross the whole genome

depth_box <- ggplot(sample_qc, aes(x = timepoint, y = glob_depth, fill = temperature)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(text = sample), position = position_jitterdodge(jitter.width = 0.1), size = 2) +
  theme_minimal() +
  labs(x = "Life stage", y = "Depth")

print(depth_box)

ggplotly(depth_box)


```

Upon looking at the really, really low samples for coverage and depth on the final day, it looks like the first spawning blocks had the lowest; this makes sense because they had the lowest pooled sample sizes. Therefore, algae contamination would be strongest with the lowest pool size...

#### sample coverage vs. depth
* cov on y axis
* depth on x axis
* sort in ascending order
* color by temperature treatment
* lines for QC

```{r sample coverage vs. depth}
cov_depth <- ggplot(sample_qc, aes(x = global_coverage, y = glob_depth, fill = temperature, color = temperature, text = sample)) +
  geom_point() +
  labs(x = "Average sample depth", y = "Average sample coverage")
  

print(cov_depth)
ggplotly(cov_depth)


```
