---
title: "coverage_and_mapping.Rmd"
author: "Emily Dombrowski"
date: "2026-07-16"
output: html_document
---

This is a R markdown file where I will be looking at mapping stats and coverage for Whole Genome Sequenced *Mytilus* mussel data. 

I have previously visualized mapping and coverage, but I had to come back because I applied some more stringent filtering to my samples. So, here are my next few analyses. 

### Load libraries and import data

```{r}

library(tidyverse)
library(plotly)

# samtools coverage data outputs
cov_data <- read_csv("master_qc_and_coverage.csv")

head(cov_data)
```

The data is set up with the following structure: 
* sample: sample name
  * either larval samples with spawning block, replicate-temperature-date-elution
  *or, adule mussel: sex, number, elution
* chromosome scaffold name
* start position of reads
* end position of reads
* total number of reads
* covbases: number of base pairs along chromosome that have at least one read per site
* breadth of coverage:
* depth:
* meamapq: mean map quality
* percentage_gc

Next, I am going to do some manipulations to assign filters to each variable that separate them by temperature and life stage.
Additionally, there are some names that don't follow the strict pattern, so I included code to manually reassign these samples. 

#### manipulate data to add filters
```{r}
cov_filter <- cov_data %>%
  mutate(timepoint = case_when(
      # Looks for strings starting with F or M
      str_detect(sample, "^[FM]")       ~ "adult",
      
      # Matches exact day tags within the sample ID
      str_detect(sample, "-D1-")          ~ "day_1",
      str_detect(sample, "-D10-")         ~ "day_10",
      
      # The | operator means "OR", catching both Day 19 and Day 20
      str_detect(sample, "-D19-|-D20-")   ~ "final_day",
      
      # Catch-all safety net just in case a sample ID is misformatted
      TRUE                                ~ "Unknown" 
    ),
    
    # 2. Extract the Temperature component
    temperature = case_when(
      str_detect(sample, "-14C-") ~ "14C",
      str_detect(sample, "-22C-") ~ "22C",
      TRUE                        ~ "initial" # Adults and Day 1 baselines land here
    )
    
  )
  
cov_filter <- cov_filter %>%
  mutate(
    temperature = case_when(
      sample == "C3R3-D20-14-E1" ~ "14C",
      sample == "B1R3-B2R1-22-D10-E1" ~ "22C",
      TRUE ~ temperature
                          )
  )
```

#### organization: total mapped reads per chromosome
```{r}
# Extract total mapped reads per sample from the "all" row
total_mapped_df <- cov_filter %>%
  filter(chromosome == "all") %>%
  select(sample, total_mapped_reads = numreads)

# Join it back to calculate proportions for individual chromosomes
cov_filter <- cov_filter %>%
  left_join(total_mapped_df, by = "sample") %>%
  mutate(
    # What proportion of the sample's total mapped reads landed on this chromosome?
    prop_reads_on_chrom = numreads / total_mapped_reads
  )
```

#### summary stats
```{r}

# Generate a summary table of overall depth and breadth
coverage_summary_table <- cov_filter %>%
  filter(chromosome == "all") %>%
  group_by(timepoint, temperature) %>%
  summarise(
    sample_count = n(),
    mean_depth = mean(depth),
    sd_depth = sd(depth),
    min_depth = min(depth),
    max_depth = max(depth),
    mean_breadth_pct = mean(breadth_of_coverage),
    .groups = "drop"
  )

```

```{r}
# Barplot showing sample vs. proportion mapped
p <- ggplot(, aes(x = sample, y = Proportion_Mapped, fill = temperature)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank()) +
  labs(y = "Mapped reads / total reads",
       x = "Sample (hover over bar to see individuals)")
p

ggplotly(p)

```

