---
title: "mapping_proportions"
author: "Emily Dombrowski"
date: "2026-07-03"
output: html_document
---

This is a script for parsing mapping stats in mussel WGS samples. The purpose of this script is to take quality control stats from the mapping process of fastq files to .sam (or in my case .cram) files and deduplication using Picard.

Part of this script crashed; so I will start a new one with the newer obtained data.
Input documents: `mapping_proportions_summary.txt` 
Output figures: 
* Coverage / sample (interactive plotly)
* 

#### Load libraries and documents
```{r}

# Load libraries
library(tidyverse)
library(plotly)

# 1. Load the mapping data (tab-separated)
qc_data <- read_delim("mapping_proportions_summary.txt", delim = "\t")
```

The names in this document are confusing, so here are some adjustments that take into account the math. For example, since there are pretty low mapping rates, the script had a step that subtracted unmapped reads from mapped reads; some of these were coming out negative, and had to be deleted. 
```{r}
qc_data <- qc_data %>%
  mutate(
    # The original 'Total_Reads' column actually represents the true mapped reads
    True_Mapped = Total_Reads, 
    True_Total  = Total_Reads + Unmapped_Reads,
    
    # Recalculate everything accurately
    Proportion_Mapped = True_Mapped / True_Total,
    Total_Reads       = True_Total,
    Mapped_Reads      = True_Mapped
  ) %>%
  dplyr::select(-True_Mapped, -True_Total)

# View the corrected structure
head(qc_data)
glimpse(qc_data)

# 2. Global Summary Statistics
summary_metrics <- qc_data %>%
  summarise(
    total_samples      = n(),
    mean_total_reads   = mean(Total_Reads),
    median_total_reads = median(Total_Reads),
    mean_mapping_prop  = mean(Proportion_Mapped),
    median_mapping_prop = median(Proportion_Mapped),
    min_mapping_prop   = min(Proportion_Mapped),
    max_mapping_prop   = max(Proportion_Mapped)
  )

print("--- Global Alignment Metrics Summary ---")
print(summary_metrics)


# 3. Identify Outlier Pools (Flagging low mapping or low read counts)
# Adjust these thresholds based on your specific sequencing depth expectations
mapping_threshold <- 0.70  # Flag pools where less than 70% of reads mapped
read_count_threshold <- 5000000  # Flag pools with fewer than 5M total reads

low_mapping_samples <- qc_data %>% 
  filter(Proportion_Mapped < mapping_threshold)

View(low_mapping_samples)

low_depth_samples <- qc_data %>% 
  filter(Total_Reads < read_count_threshold)

print(paste("Samples below mapping threshold:", nrow(low_mapping_samples)))
print(low_mapping_samples)

print(paste("Samples below depth threshold:", nrow(low_depth_samples)))
print(low_depth_samples)

# 4. Visualization

# Plot A: Distribution of Mapping Proportions
ggplot(qc_data, aes(x = Proportion_Mapped)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Mytilus Pool Mapping Success",
    x = "Proportion of Reads Mapped",
    y = "Number of Samples"
  )

# Plot B: Total Depth vs. Mapping Success
q <- ggplot(qc_data, aes(x = Total_Reads / 1e6, y = Proportion_Mapped)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Sequencing Depth vs. Alignment Efficiency",
    x = "Total Reads (Millions)",
    y = "Proportion Mapped"
  )

ggplotly(q)

```


```{r coverage histogram}

# Barplot showing sample vs. proportion mapped
p <- ggplot(qc_data, aes(x=Sample, y=Proportion_Mapped)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank()) +
  labs(y = "Mapped reads / total reads",
       x = "Sample (hover over bar to see individuals)")

ggplotly(p)

# Total reads per sample
m <- ggplot(qc_data, aes(x=Sample, y=Mapped_Reads)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank()) +
  labs(y = "Mapped reads",
       x = "Sample")

ggplotly(m)

```

```{r subset by time point and temperature}

qc_subset <- qc_data %>%
  mutate(
    Timepoint = case_when(
      # Looks for strings starting with F or M
      str_detect(Sample, "^[FM]")       ~ "adult",
      
      # Matches exact day tags within the sample ID
      str_detect(Sample, "-D1-")          ~ "day_1",
      str_detect(Sample, "-D10-")         ~ "day_10",
      
      # The | operator means "OR", catching both Day 19 and Day 20
      str_detect(Sample, "-D19-|-D20-")   ~ "final_day",
      
      # Catch-all safety net just in case a sample ID is misformatted
      TRUE                                ~ "Unknown" 
    ),
    
    # 2. Extract the Temperature component
    Temperature = case_when(
      str_detect(Sample, "-14C-") ~ "14C",
      str_detect(Sample, "-22C-") ~ "22C",
      TRUE                        ~ "initial" # Adults and Day 1 baselines land here
    )
  )

# two samples with bad naming conventions; let's manually reassign them to the correct category
# here are the sample names: 
# C3R3-D20-14-E1
# B1R3-B2R1-22-D10-E1

qc_subset <- qc_subset %>%
  mutate(
    Temperature = case_when(
      Sample == "C3R3-D20-14-E1" ~ "14C",
      Sample == "B1R3-B2R1-22-D10-E1" ~ "22C",
      TRUE ~ Temperature
                          )
        )

# Boxplot to visualize the interaction of time and temperature
box <- ggplot(qc_subset, aes(x = Timepoint, y = Proportion_Mapped, fill = Temperature)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1), size = 2) +
  theme_minimal() +
  labs(x = "Life stage", y = "Mapped reads/unmapped reads")

ggplotly(box)

print(box)

```